### 2022-02-04 article.tex
Performance-Based Evaluation
Generalizing Scala's Singleton Types

We show the type safety of \SystemFm through the usual progress and preservation theorems.
This section provides an overview of the proof structure and states the involved lemmas and theorems.
Detailed proofs are available in the supplementary material of this paper, in two different versions.
The first version, \cite{blanvillain2021type}, is a pen-and-paper proof where \SystemFm is exactly as presented in \Cref{fig:FMTypingRules}.
The second version, \cite{blanvillain2021artifact}, is a mechanization of the proof in Coq,
using the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings.
Our mechanization uses a simplified representation of match types with exactly one case per match.
Matches with multiple cases can be expressed by nesting match types in default cases.

@@

\subsection{Caching}

@@

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014understanding}.
Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.

>>>>

We show the type safety of \SystemFm through the usual progress and preservation theorems.
This section provides an overview of the proof structure and states the involved lemmas and theorems.
Detailed proofs are available in the supplementary material of this thesis, in two different versions.
The first version, \cite{blanvillain2021type}, is a pen-and-paper proof where \SystemFm is exactly as presented in \Cref{fig:FMTypingRules}.
The second version, \cite{blanvillain2021artifact}, is a mechanization of the proof in Coq,
using the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings.
Our mechanization uses a simplified representation of match types with exactly one case per match.
Matches with multiple cases can be expressed by nesting match types in default cases.

@@

\subsection{Caching}
\label{subsec:caching}

@@

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014understanding}.
Nevertheless, we believe that the results of this chapter are directly applicable to conditional types and could be used to improve TypeScript's type checker.

### 2022-02-04 preprint.tex
Generalizing Scala's Singleton Types

In this paper, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

@@

\begin{itemize}
\item
  We begin by motivating why type-level programming in desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}). Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
\item
  TODO
  We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
  We describe how we extended Scala with dependent types.
  This development was an experiment to explore the feasibility of adding dependent types in Scala.
  We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
  Our presentation focuses on several facets of the implementation that are not reflected in our formalism.

  We describe how we extended Scala with dependent types (\Cref{sec:extending-scala}). This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in adding dependent types to their language with subtyping.
\item
  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
  TODO: keep this?
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

TODO: pretend that formalization doesn't exist?
The said report covers more material than what is included in this chapter, in particular, our formalization and the associated metatheory are out of the scope of this dissertation \citep[Section 3 and 4]{schmid2020coming}.

@@

\begin{lstlisting}
dependent val schema1 = Cons("age", Cons("name", Nil()))
dependent val schema2 = Cons("name", Cons("unit", Nil()))
dependent val table1  = Table(schema1, (*\text{\dots}*))
dependent val table2  = Table(schema2, (*\text{\dots}*))
dependent val joined  = table1.join(table2, "name")
joined: { Table(
            Cons("name", Cons("age", Cons("unit", Nil()))),
            _: DataFrame) }
\end{lstlisting}

@@

In this section we give an overview of how we extended Scala with a generalized notion of singleton types, and how we manipulate these types during type-checking.
Our presentation is divided into three parts.
First, we introduce new types that reflect a subset of Scala's term language at the type level.
We then introduce the |dependent| qualifier, which influences type inference to assign generalized singleton types to the annotated definition.
Finally, we give an informal description of our type evaluation algorithm.

@@

\subsection{Reflecting terms in types}

The fundamental difference between our and Scala's existing type system is that we take Scala's notion of singleton types, that is, precise types for variable bindings and literals, and extend them to cover Scala's core functional expressions.
Concretely we add new forms of types for the following constructs:

@@

Such new types can be expressed using the |{| $e$ |}| type syntax, where $e$ is an expression in the core functional subset of Scala listed above.
For instance, suppose $\text{foo}$ stands for the expression

@@

The first two constructs in this list have antecedents in Scala. Types for bindings are available since the early days of Scala with the |x.type| syntax \citep[Section 3.2.1]{scalaSpec}, which is equivalent to |{ x }| in our system.
Types for primitive literals have recently been added to the language: literals for booleans, strings and the various numeric types are made available in the type language \citep{sip23}.

@@

\begin{itemize}
  \item |Cons(Nil, Nil): { Cons(_: Any, Nil   ) }|
  \item |Cons(Nil, Nil): { Cons(Nil   , _: Lst) }|
  \item |Cons(Nil, Nil): { Cons(_: Any, _: Lst) }|
\end{itemize}

@@

@@

\noindent
When used in a type, this expression reduces to |v1| if the scrutinee's type is a subtype of |T1|.
In order to reduce to |v2|, type normalization must make sure |T1| and the scrutinee's type are disjoint, namely that the dynamic type of |s| cannot possibly be smaller than |T1|.
Disjointness proofs are built using static knowledge about the class hierarchy and make use of the guarantees implied by the |sealed| and |final| qualifiers, which are Scala's way of declaring closed-type hierarchies.

@@

@@

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Thanks to |choose[T]| we can still formulate precise result types when terms depend on the result of side-effectful operations.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type, but we can still assign a better type than |Lst| to an expression like |Cons(z, Nil())|, that is, |{ Cons(choose[Int], Nil()) }|.

@@

Similarly to how we model other side effects, exceptions are approximated in types.
Our type-inference algorithm uses a new error type, |Error(e)|, which we infer when raising an exception with |throw e|.
Exception handlers are typed imprecisely using the default mode of type-inference.
Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{e1; e2}| is simply |{ e2 }|.
However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.
In these cases, our type system can prove that the program execution will encounter exceptional behavior, and reports a compilation error.
This approach is conservative in that it might reject programs that recover from exceptions.
Also note that this is a sanity check, rather than a guarantee of no exceptions occurring at runtime.
That is, depending on which rules are used during subtyping, it is possible to succeed without entering type normalization, resulting in such errors going undetected.
Despite these shortcomings, our treatment of exceptions results in a practical way to raise compile-time errors.
It would be interesting to explore the addition of an effect system to our Scala extension and formalization.

@@

The first question is whether type-checked programs are guaranteed to terminate.
For simplicity, our work side-steps this question, requiring bounds for recursion.
A more general solution would be to compute or infer such bounds using measure functions, as done in \FR~\citep{hamza2019system}.
Another approach would be to extend our translation of non-determinism to permit non-termination.
We consider this aspect orthogonal to the objectives of this paper.
Our work targets general-purpose programming language whose type safety is defined with regards to its runtime semantics and that may include non-terminating interactive computations.

The second question is termination of our type checker.
Non-termination of type checking implies that the type checker can give three possible answers, "type correct", "type incorrect" or "do not know" (or timeout).
Treating "do not know" as "type incorrect" makes the non-termination unproblematic from a soundness perspective.
A similar argument is made for other dependently-typed languages with unbounded recursion, such as Dependent Haskell~\citep{eisenberg2016dependent} or Cayenne~\citep{augustsson1998cayenne}.
In practice, our system deals with infinite loops using a fuel mechanism.
Every evaluation step consumes a unit of fuel, and an error is reported when the compiler runs out of fuel.
The default fuel limit can be increased via a compiler flag to enable arbitrarily long compilation times.

>>>>

In this chapter, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

@@

\begin{itemize}
\item
  We begin by motivating why type-level programming in desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}). Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
\item
  We describe how we extended Scala with a generalization of singleton types.
  We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
  This practical introduction would be of interest to any Scala programmer willing to learn how to use our system,
  as well as programming language designers interested dependent types.
\item
  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

@@

\begin{lstlisting}
dependent val schema1 = Cons("age", Cons("name", Nil()))
dependent val schema2 = Cons("name", Cons("unit", Nil()))
dependent val table1  = Table(schema1, (*\text{\dots}*))
dependent val table2  = Table(schema2, (*\text{\dots}*))
dependent val joined  = table1.join(table2, "name")
joined: { Table(Cons("name", Cons("age", Cons("unit", Nil()))), _: DataFrame) }
\end{lstlisting}

@@

In this section we give an overview of how we extended Scala with a generalized notion of singleton types, and how we manipulate these types during type-checking.
Our presentation is divided into three parts.

First, we introduce new types that reflect a subset of Scala's term language at the type level, and give an informal description of our type evaluation algorithm.
We then introduce the |dependent| qualifier, which influences type inference to assign generalized singleton types to the annotated definition.
Finally, we discuss how our extension interacts with other aspects of Scala, such as side effects, virtual dispatch, and recursion.

@@

\subsection{Reflecting Terms in Types}

The fundamental difference between our and Scala's existing type system is that we take Scala's notion of singleton types, that is, precise types for variable bindings and literals, and extend them to cover Scala's core functional expressions.
Concretely we add new types for the following constructs:

@@

These new types can be expressed using the |{| $e$ |}| type syntax, where $e$ is an expression in the core functional subset of Scala listed above.
For instance, suppose $\text{foo}$ stands for the expression

@@

The first two constructs in this list have antecedents in Scala.
Types for bindings are available since the early days of Scala with the |x.type| syntax \citep[Section 3.2.1]{odersky2006scala}, which is equivalent to |{ x }| in our system.
Types for primitive literals have recently been added to the language: literals for booleans, strings and the various numeric types are made available in the type language \citep{leontiev2014sip}.

@@

\begin{itemize}
  \item |Cons(Nil, Nil): { Cons(_: Any, Nil) }|
  \item |Cons(Nil, Nil): { Cons(Nil, _: Lst) }|
  \item |Cons(Nil, Nil): { Cons(_: Any, _: Lst) }|
\end{itemize}

@@

\subsection{Type Evaluation}

Our system evaluates types using a call-by-value partial evaluator, which we embedded in the type-checker.
The evaluation rules for the term in generalized singleton types are completely standard up to dynamic type tests.

We perform evaluation of types during subtyping.
For instance, given the following definition of a |paritiy| function:

\begin{lstlisting}
dependent def parity(x: Int) =
  if (x % 2 == 0) "even" else "odd"
\end{lstlisting}

in order to prove the well-typedness of

\begin{lstlisting}
val p: { "odd" } = parity(5)
\end{lstlisting}

\noindent
the type-checker will ensure that |{ parity(5) }| is a subtype of |{ "odd" }|.
In the process, the left-hand side is evaluated as follows:

\begin{lstlisting}
  { parity(5) }                          (*$(1)$*)
= { if (5 % 2 == 0) "even" else "odd" }  (*$(2)$*)
= { if (false) "even" else "odd" }       (*$(3)$*)
= { "odd" }
\end{lstlisting}

\noindent
In (1), the method call to |parity| is replaced by its result type, where the method parameter has been substituted by the singleton type of the argument, |{ 5 }|.
(2) evaluates the boolean expression to |false|.
(3) reduces the |if| expression to its else branch.

In general, our evaluator will execute operations on concrete primitive values of types such as |Boolean|, |Int| and |String|, i.e., perform constant-folding.

@@

\noindent
When used in a type, this expression reduces to |v1| if the scrutinee's type is a subtype of |T1|.
In order to reduce to |v2|, type normalization must make sure |T1| and the scrutinee's type are disjoint, namely that the dynamic type of |s| cannot possibly be smaller than |T1|.
Disjointness proofs are built using static knowledge about the class hierarchy, such as the |sealed| and |final| qualifiers, which are Scala's way of declaring closed-type hierarchies.

@@

\noindent
Note that we could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type  explicitly.

@@

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type, since reading a mutable variable is a side effect.
However, we can still assign precise types to expressions containing |z|, for instance, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.

@@

Similarly to how we model other side effects, exceptions are approximated in types.
Our type-inference algorithm uses a new error type, |Error(e)|, which we infer when raising an exception with |throw e|.
Exception handlers are typed imprecisely using the default mode of type-inference.
Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{e1; e2}| is simply |{ e2 }|.
However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.
In these cases, our type system can prove that the program execution will encounter exceptional behavior, and reports a compilation error.
This approach is conservative in that it might reject programs that recover from exceptions.
Also note that this is a sanity check, rather than a guarantee of no exceptions occurring at runtime.
That is, depending on which rules are used during subtyping, it is possible to succeed without entering type normalization, resulting in such errors going undetected.
Despite these shortcomings, our treatment of exceptions results in a practical way to raise compile-time errors.
It would be interesting to explore the addition of an effect system to our Scala extension.

@@

The first question is whether type-checked programs are guaranteed to terminate.
Proving termination or totality of dependent programs is a non-goal of our system.
Unlike proof assistants, Scala programs do not manipulate proof terms.
Consequently, the lack of totality checks does not affect Scala's present notion of safety.
Exceptions or infinite loops in the evaluation of a `dependent` method would prevent the completion of type-checking.

The second question is termination of our type checker.
Non-termination of type checking implies that the type checker can give three possible answers, "type correct", "type incorrect" or "do not know" (when type checking times out).
Treating "do not know" as "type incorrect" makes the non-termination unproblematic from a soundness perspective.
A similar argument is made for other dependently-typed languages with unbounded recursion, such as Dependent Haskell~\citep{eisenberg2016dependent} or Cayenne~\citep{augustsson1998cayenne}.
In practice, our system deals with infinite loops using a fuel mechanism.
Every evaluation step consumes a unit of fuel, and an error is reported when the compiler runs out of fuel.
The default fuel limit can be increased via a compiler flag to enable arbitrarily long compilation times.
